---
title: "Soil Carbon data validation"
date: "2025-01-27"
output: html_document
---

```{r setup, include=FALSE}
# Load libraries, WD, cran repository
library(DBI)
library(RSQLite) 
library(dplyr)
library(geosphere)
library(sp)
library(ggplot2)
library(rnaturalearth)
library(sf)
library(maps)
library(scoringRules)
library(ncf)
library(terra)
library(knitr)
library(gstat)
library(tidyr)
library(neonstore)
library(neonUtilities)
library(swfscMisc)
library(PEcAn.utils)
library(maps)


# load functions from other script
source("extract_neon_sc_func.R")

setwd("/Users/chane/OneDrive/Documents/ef lab/R files")
options(repos = c(CRAN = "https://cran.rstudio.com/"))
```


Get soil carbon data from SDA forecast (copying Cami's code)
```{r, echo = FALSE}

# Load data
SIPNET <- load("sda.all.forecast.analysis.Rdata")

# Get site coordinates
Site_Info <- readRDS("site.locs.rds")

# Rename lists to years
names(analysis.all) <- 2012:2021
year <- 2012:2021

# Names of variables
variables <- c("AbvGrndWood", "LAI", "SoilMoistFrac", "TotSoilCarb")

# Loop through each year list in analysis.all
for (year in names(analysis.all)) {
  
  # Get the current data frame for the year
  year_df <- analysis.all[[year]]
  
  # Initialize a vector for new column names
  new_colnames <- character(ncol(year_df))
  
  # Loop over each site and assign variable names
  for (site in 1:6400) {
    # Calculate the starting column index for each site's variables
    start_col <- (site - 1) * 4 + 1
    
    # Assign the names for the 4 variable columns for this site
    new_colnames[start_col:(start_col + 3)] <- paste0(variables, "_Site", site)
  }
  
  # Apply the new column names to the data frame
  colnames(year_df) <- new_colnames
  analysis.all[[year]] <- year_df  # Update the list with modified data frame
}


# Load list of neon sites with their latitudes and longitudes
Neonplots <- read.csv("Neon_sites_terrestrial.csv") 

# Add a column for site numbers in Site_Info (site_info has coordinates of each location in df)
Site_Info$Site_Number <- seq(1, nrow(Site_Info)) 

  # Extract latitude and longitude from Neonplots and Site_Info tables
neon_coords <- Neonplots[ ,2:3] %>%
  select(Latitude = field_latitude, Longitude = field_longitude)
Site_Info_coords <- Site_Info %>%
  select(Site_Number, lat, lon)

  # Put Site_Info latitude/longitude in a matrix
Site_Coordinates <- as.matrix(Site_Info[, c("lat", "lon")])

  # Same for Neonplots
Neonplots_Coordinates <- as.matrix(Neonplots[, 2:3])

  # Create a new column in Neonplots to store the closest Site_Number
Neonplots$Closest_Site <- NA

# Loop through each row in Neonplots and calculate each plot's closest site number
for (i in 1:nrow(Neonplots)) {
  # Extract the coordinates for the current row in Neonplots
  plot_coords <- Neonplots_Coordinates[i, ]
  
  # Compute distances from this plot to all site coordinates (longlat means its in km)
  distances <- spDistsN1(Site_Coordinates, plot_coords, longlat = TRUE)
  
  # Find the index of the minimum distance for closest site
  closest_site_index <- which.min(distances)
  
  # Assign the corresponding Site_Number from Site_Info to the current row in Neonplots
  Neonplots$Closest_Site[i] <- Site_Info$Site_Number[closest_site_index]}

```

Filtering analysis data with relevant sites
```{r, echo=FALSE}
## Filter analysis.all to only include relevant sites ##

# Get the list of sites to keep from Neonplots
sites_to_keep <- Neonplots$Closest_Site  

# Loop through each data frame in the analysis.all list
analysis.fia = list()
for (year in names(analysis.all)) {
  # Get the current data frame
  df <- analysis.all[[year]]
  
  # Create a pattern to match columns that include sites in sites_to_keep
  pattern <- paste0("_Site(", paste(sites_to_keep, collapse = "|"), ")$")
  
  # Keep only columns that match the pattern
  df_filtered <- df[, grep(pattern, colnames(df))]
  
  # Replace the original data frame in the list with the filtered one
  analysis.fia[[year]] <- df_filtered}

```

Now extract the sites from Neonplots
```{r, echo = FALSE}
## Extract the unique site-year combinations from Neonplots ##
valid_sites <- unique(Neonplots[, "Closest_Site"])

# Loop over each year list in analysis.all
for (year in names(analysis.fia)) {
  
  # Extract the current data frame for the year
  year_df <- analysis.fia[[year]]
  
  # Identify the year as a numeric value
  current_year <- as.numeric(substr(year, 1, 4))
  
  # Identify columns with site numbers that match valid_sites
  matching_columns <- grep(paste0("_Site(", paste(valid_sites, collapse = "|"), ")$"), 
                           colnames(year_df), value = TRUE)
  
  # Subset the data frame to only include matching columns
  analysis.fia[[year]] <- year_df[, matching_columns, drop = FALSE]
}
```

Extract soil carbon data - mean and standard deviation of each year
```{r, echo = FALSE}
# Initialize a list to store summary tables for each year
summary_list <- list()

# Loop through each year's data frame in analysis.all
names(analysis.fia) = names(analysis.all)
for (year in names(analysis.all)) {
  
  # Extract the year from the full date format 
  year_only <- sub("^(\\d{4}).*", "\\1", year)
  
  # Get the current data frame
  df <- analysis.fia[[year]]
  
  # Create a pattern to match columns that include sites
  site_numbers <- unique(gsub(".*_Site(\\d+)", "\\1", colnames(df))) 
  
  # Initialize a data frame to hold summary statistics for this year
  summary_table <- data.frame(Site_Number = numeric(0), 
      Year = character(0), 
      SC_Mean_Mod = numeric(0), 
      SC_SD_Mod = numeric(0),
      stringsAsFactors = FALSE)

  # Loop through each site number
  for (site in site_numbers) {
    # Create a vector for soil carbon data
    tsc_col <- paste0("TotSoilCarb_Site", site)
    
     # Check if columns exist in the data frame
    if (tsc_col %in% colnames(df)) {
       soilC = df[[tsc_col]]
       
      # Calculate mean and SD for each variable
      mean_soil = mean(soilC, na.rm = TRUE)  # Average of soil C
      sd_soil = sd(soilC, na.rm = TRUE)      # Standard deviation of soil C
      
      # Add to summary table
      summary_table <- rbind(summary_table, 
        data.frame(Site_Number = as.numeric(site), 
        Year = year_only,  
        SC_Mean_Mod = mean_soil,
        SC_SD_Mod = sd_soil))
    }
  }
 # Store the summary table for this year in the list
  summary_list[[year]] <- summary_table 
}  

# Combine all the summary tables into one data frame
final_summary_table <- do.call(rbind, summary_list) 
```

Graph the soil carbon year means in a histogram 
```{r}
hist(final_summary_table[["SC_Mean_Mod"]], main = "Histogram of Soil Carbon Means across NEON Sites and Year", col = "lightblue", breaks = 10, border = "black", freq = FALSE)
```


Loading soil carbon data from NEON (Alexi's code)
```{r, echo = FALSE}

# load copy of Alexi's code
source("extract_neon_sc_func.R") 

# Set start and end to NA to get all years
start_date = NA
end_date = NA 
store_dir <- "/Users/chane/OneDrive/Documents/ef_lab/R_files" # storing to folder I am in

# create summary table for loading data 
summary_table_NEON_data <- data.frame(sitename = character(240),
            year = character(240), 
           mean_organicC = numeric(240),
            sd_organicC = numeric(240),
            mean_bulkD = numeric(240),
            sd_bulkD = numeric(240),
            mean_frac30 = numeric(240),
            sd_frac30 = numeric(240),
            mean_soilC= numeric(240),
            sd_soilC = numeric(240))

# create temp data 
temp_table <- data.frame(sitename = character(240),
            year = character(240), 
            mean_organicC = numeric(240),
            sd_organicC = numeric(240),
            mean_bulkD = numeric(240),
            sd_bulkD = numeric(240),
            mean_frac30 = numeric(240),
            sd_frac30 = numeric(240),
            mean_soilC= numeric(240),
            sd_soilC = numeric(240))

# assign first data to the table
sitename <- Neonplots["field_site_id"][1,1]
summary_table_NEON_data <-extract_NEON_veg(sitename, start_date, end_date, store_dir, neonsites = NULL)

# loop through and mean and sd of each data point from each year at all of the neonplot sites
for (i in 2:47){
  sitename <- Neonplots["field_site_id"][i,1]
  temp_table <- extract_NEON_veg(sitename, start_date, end_date, store_dir, neonsites = NULL)
  summary_table_NEON_data <- rbind(summary_table_NEON_data, temp_table)
}

temp_summary_NEON <- summary_table_NEON_data

```

Add model to summary table
```{r}
# create summary_table as copy of neon table
summary_table <- temp_summary_NEON

# remove NAs
summary_table <- drop_na(summary_table)

# add row for sitename
summary_table$Site_Number <- numeric(length(summary_table$sitename)) 


# loop through summary table and Neon plots site and when equal and add closest site
for (i in 1:length(summary_table$year)){
   for (j in 1:length(Neonplots$field_site_id)){
      if (Neonplots$field_site_id[j] == summary_table$sitename[i]){
          summary_table$Site_Number[i] <- Neonplots$Closest_Site[j]
   }
 }
}

# sort by site number
summary_table <- summary_table[order(summary_table$Site_Number),]

# sort based on year (like final_summary_table)
summary_table <- summary_table[order(summary_table$year),]

# add rows for forecast to the summary_table_NEON_data
summary_table$SC_mean_forecast = numeric(nrow(summary_table));
summary_table$SC_sd_forecast = numeric(nrow(summary_table))

c = 1 # counter

# add forecast data, leave zeros where no forecast data
for (i in 1:length(final_summary_table$Year)){
  for (j in 1:nrow(summary_table)){
    if (summary_table$year[j] < 2022){
      if (final_summary_table$Year[i] == summary_table$year[j] & final_summary_table$Site_Number[i] == summary_table$Site_Number[j]){
        summary_table$SC_mean_forecast[c] = final_summary_table$SC_Mean_Mod[i];
        summary_table$SC_sd_forecast[c] = final_summary_table$SC_SD_Mod[i];
        c <- c +1
      }
    }
  }
}


```

Analysis!
```{r, echo = FALSE}
# Plot summary table

summary_table_2021 = summary_table[1:77, 1:12]
ggplot(summary_table_2021, aes(x=SC_mean_forecast, y=mean_soilC)) + geom_point(size = 2)

```

Map of RMS error
```{r}
# find the error 
summary_table$error_soilC = numeric(nrow(summary_table))

for (i in 1:nrow(summary_table)){
  # model - actual
  summary_table$error_soilC[i] = (summary_table$SC_mean_forecast[i] - summary_table$mean_soilC[i])
} 

# find rms error for each site

# make dataframe to store name and rms error of each
error_df <- data.frame(sitename = character(46),
            rms_error = numeric(46))

error_df$sitename <- unique(summary_table$sitename) # make list of names

# create temp vector for before averages
vec <- numeric(1)
c = 1 # counter

for (i in 1:nrow(error_df)){
  for (j in 1:nrow(summary_table)){
    if(error_df$sitename[i] == summary_table$sitename[j]){
      vec[c] = (summary_table$error_soilC[j])^2
      c <- c + 1
    }
  }
  error_df$rms_error[i] = sqrt(sum(vec)/(c-1))
  c <- 1
  rm(vec)
  vec <- numeric(1)
}

# sort alphabetically
error_df <- error_df[order(error_df$sitename),]

# add lat and long columns
error_df$site_lat <- numeric(nrow(error_df))
error_df$site_long <- numeric(nrow(error_df))

# add lat and long
for (i in 1:nrow(error_df)){
  for (j in 1:nrow(Neonplots)){
    if (error_df$sitename[i] == Neonplots$field_site_id[j]){
      error_df$site_lat[i] <- Neonplots$field_latitude[j]
      error_df$site_long[i] <- Neonplots$field_longitude[j]
    }
  }
}

# get map of usa
USA = map_data("world") %>% filter(region == "USA")

# make bubble map with size and color depening on RMS error
ggplot()+
  geom_polygon(data = USA, aes(x= long, y = lat, group = group), fill="forestgreen", alpha = 0.3, main = "RMS  error of each site") + geom_point(data = error_df, aes(x= site_long, y = site_lat, size = rms_error, color = rms_error)) + geom_point(alpha = 0.7) + xlim(-170, -50) + scale_size_continuous(range = c(1,10)) + theme_void() + labs(title = "RMS error of forecast for NEON sites") + theme(plot.title = element_text(hjust = 0.5),)

# find max errors
error_df <- error_df[order(error_df$rms_error),]

```

Find rate of change of soil carbon for each 
```{r, echo = FALSE}

# find rate of change of NEON data
# sort by location
summary_table <- summary_table[order(summary_table$sitename),]

error_df <- error_df[order(error_df$sitename),]

# create places for rate of change
error_df$roc_soilC <- numeric(46)
error_df$roc_forecast <- numeric(46)
error_df$roc_percent_error <- numeric(46)
error_df$roc_error <- numeric(46)

c = 1
for (i in 1:nrow(summary_table)){
  if (i == 1){
    if( summary_table$sitename[1] != summary_table$sitename[2]){
      error_df$roc_soilC[1] <- NaN
      error_df$roc_forecast[1] <- NaN
      error_df$roc_percent_error[1] <- NaN
      error_df$roc_error <- NaN
    }
  }
  if (i>1){
    if (summary_table$sitename[i] != summary_table$sitename[i - 1]){
      temp_name = summary_table$sitename[i]
      j <- i
      while (summary_table$sitename[j] == temp_name){
        j <- j+1
        if (j == 108){
          break
      }
    }
     c <- c+1
  error_df$roc_soilC[c] <- ((summary_table$mean_soilC[j-1] - summary_table$mean_soilC[i])/(as.numeric(summary_table$year[j-1]) - as.numeric(summary_table$year[i])))
  error_df$roc_forecast[c] <- ((summary_table$SC_mean_forecast[j-1] - summary_table$SC_mean_forecast[i])/(as.numeric(summary_table$year[j-1]) - as.numeric(summary_table$year[i])))
  error_df$roc_error[c] <- (error_df$roc_soilC[c] - error_df$roc_forecast[c])
  error_df$roc_percent_error[c] <- abs((error_df$roc_soilC[c] - error_df$roc_forecast[c])/error_df$roc_forecast[c])*100
  }
  }
}

# Bubble map with size and color depending on RMS error
ggplot()+
  geom_polygon(data = USA, aes(x= long, y = lat, group = group), fill="forestgreen", alpha = 0.3) + geom_point(data = error_df, aes(x= site_long, y = site_lat, size = roc_error, color = roc_error)) + geom_point(alpha = 0.7) + xlim(-170, -50) + scale_size_continuous(range = c(1,10)) + theme_void() + labs(title = "Rate of Change Residual") + theme(plot.title = element_text(hjust = 0.5),)

ggplot()+
  geom_polygon(data = USA, aes(x= long, y = lat, group = group), fill="forestgreen", alpha = 0.3) + geom_point(data = error_df, aes(x= site_long, y = site_lat, size = roc_percent_error, color = roc_percent_error)) + geom_point(alpha = 0.7) + xlim(-170, -50) + scale_size_continuous(range = c(1,10)) + theme_void() + labs(title="Rate of Change Percent Error") + theme(plot.title = element_text(hjust = 0.5),)

error_df <- error_df[order(error_df$roc_percent_error),]

error_df_roc <- drop_na(error_df)
```

Residual map of rate of change (copied from Cami's code)
```{r, echo = FALSE}

# define colors based on sign
error_df$Color <- with(error_df,
  ifelse(roc_error > 0, "lightsalmon",
  ifelse(roc_error < 0, "orangered2", "grey")))


USA = map_data("world") %>% filter(region == "USA")
us_map <- st_as_sf(USA, 
            coords = c("long", "lat"), 
            crs = 4326)  # CRS 4326 is WGS84 (latitude/longitude)
                                

# Make the plot
ggplot(data = us_map) +
  geom_sf(fill = "lightgrey") +  
  geom_point(data = error_df, 
             aes(x = site_long, y = site_lat, color = Color, size = abs(roc_error)), 
             alpha = 0.7) +
  scale_color_manual(
    name = "Residuals",
    values = c(
      "lightsalmon" = "lightsalmon", 
      "orangered2" = "orangered2"
    ),
    labels = c(
      "lightsalmon" = "Rate of change (+)",
      "orangered2" = "Rate of change (-)"
    )
  ) +
  scale_size(range = c(1, 10), name = "Residual Size") +
  coord_sf(xlim = c(-160, -60), ylim = c(25, 70)) +
  labs(title = "Soil Carbon rate of change1
       Residuals", x = "Longitude", y = "Latitude") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5), 
  )


```

spatial variograms (from cami's code)
```{r, echo = FALSE}

proj4string(error_df) <- CRS("+proj=longlat +datum=WGS84")

variogram_model_rmse <- variogram(rms_error ~ 1, data = error_df)

ggplot(data = variogram_model_rmse, aes(x = dist, y = gamma)) +
  geom_point() +
  geom_line() +
  labs(title = "Variogram of RMS Error",
       x = "Distance",
       y = "Semivariance (γ)") +
  theme_minimal() +
  theme(
  plot.title = element_text(hjust = 0.5, margin = margin(b = 20)),
    axis.title.x = element_text(margin = margin(t = 10)), 
    axis.title.y = element_text(margin = margin(r = 10)))


```

```{r, echo = FALSE}
# coordinates(error_df_roc) <- ~ site_long + site_lat
variogram_model_roc <- variogram(roc_error ~ 1, data = error_df_roc)

ggplot(data = variogram_model_roc, aes(x = dist, y = gamma)) +
  geom_point() +
  geom_line() +
  labs(title = "Variogram of Rate of Change residual",
       x = "Distance",
       y = "Semivariance (γ)") +
  theme_minimal() +
  theme(
  plot.title = element_text(hjust = 0.5, margin = margin(b = 20)),
    axis.title.x = element_text(margin = margin(t = 10)), 
    axis.title.y = element_text(margin = margin(r = 10)))

ggplot(data = us_map) +
  ggplot
```

Compare error to 13 other factors (from Dongchen's folder on github)
```{r, echo = FALSE}

# get data on US 13 factors
dat = terra::rast("all_data_layers.tif")

# add columns to error_df
error_df$ysd <- numeric(46)
error_df$abg<- numeric(46)
error_df$fia <- numeric(46)
error_df$gedi<- numeric(46)
error_df$twi<- numeric(46)
error_df$tavg <- numeric(46)
error_df$srad <- numeric(46)
error_df$prec <- numeric(46)
error_df$vapr <- numeric(46)
error_df$ph <- numeric(46)
error_df$n <- numeric(46)
error_df$soc <- numeric(46)
error_df$sand <- numeric(46)

pt = numeric(1)

# extract data for each from each Neon site coordinate
for (i in 1:nrow(error_df)){
  pt = terra::extract(dat,data.frame(lon=error_df$site_long[i], lat = error_df$site_lat[i]))
  error_df$ysd[i] = pt$year_since_disturb
  error_df$abg[i] <- pt$agb
  error_df$fia[i] <- pt$fia
  error_df$gedi[i] <- pt$gedi
  error_df$twi[i] <- pt$twi
  error_df$tavg[i] <- pt$tavg
  error_df$srad[i] <- pt$srad
  error_df$prec[i] <- pt$prec
  error_df$vapr[i] <- pt$vapr
  error_df$ph[i] <- pt$PH
  error_df$n[i] <- pt$N
  error_df$soc[i] <- pt$SOC
  error_df$sand[i] <- pt$Sand
}

# use lm to fit a linear model to the data
fit_ysd = lm(error_df$ysd ~ error_df$rms_error)
fit_abg = lm(error_df$abg ~ error_df$rms_error)
fit_fia = lm(error_df$fia ~ error_df$rms_error)
fit_gedi = lm(error_df$gedi ~ error_df$rms_error)
fit_twi = lm(error_df$twi ~ error_df$rms_error)
fit_tavg = lm(error_df$tavg ~ error_df$rms_error)
fit_srad = lm(error_df$srad ~ error_df$rms_error)
fit_prec = lm(error_df$prec~ error_df$rms_error)
fit_vapr = lm(error_df$vapr ~ error_df$rms_error)
fit_ph = lm(error_df$ph ~ error_df$rms_error)
fit_n = lm(error_df$n ~ error_df$rms_error)
fit_soc = lm(error_df$soc ~ error_df$rms_error)
fit_sand = lm(error_df$sand ~ error_df$rms_error)


# make data frame
summary_factors <- data.frame(factor_name= character(13),
                    r_squared = numeric(13),
                    slope = numeric (13),
                    p_value = numeric(13))

for (i in 23:35){
  # fit a line
  error_df$temp = error_df[[i]]
  fit = lm(error_df$temp ~ error_df$rms_error)
  
  # graph

  print(ggplot(error_df, aes(x = temp, y = rms_error)) + geom_point() + geom_smooth(method = "lm", se = FALSE) + xlab(colnames(error_df[(i-13)])))

  # add r squared, slope, p value
  summary_factors$factor_name[(i-22)] <- colnames(error_df[(i-13)])
  summary_factors$slope[(i-22)] <- coef(fit)[2]
  f <- summary(fit)$fstatistic
  summary_factors$p_value[(i-22)] <- pf(f[1], f[2], f[3], lower.tail=F)
  summary_factors$r_squared[(i-22)] <- summary(fit)$adj.r.squared
}
```







