---
title: "Soil Resp"
output: html_document
date: "2025-04-22"
---

Libraries
```{r setup, include=FALSE}
library(DBI)
library(RSQLite) 
library(dplyr)
library(geosphere)
library(sp)
library(ggplot2)
library(sf)
library(maps)
library(terra)
library(knitr)
library(tidyr)
library(neonUtilities)
library(swfscMisc)
library(maps)
library(tidyverse)
library(neonSoilFlux)
library(parallel)
library(ncdf4)
library(data.table)
library(purrr)
library(data.table)
library(parallel)
library(lubridate)
library(terra)
library(httr)
library(sp)
library(neonSoilFlux)
library(stringr)
library(scales)
```

Get NEON soil respiration data
```{r}
# get a list of NEON sites and their coordinates
Neonplots <- read.csv("/usr4/ugrad/chaney/R/3_25/Neon_sites_terrestrial.csv") 

# Add a column for site numbers in Site_Info (site_info has coordinates of each location in df)
Site_Info <- readRDS("/usr4/ugrad/chaney/R/3_25/site.locs.rds")

Site_Info$Site_Number <- seq(1, nrow(Site_Info)) 

# Extract latitude and longitude from Neonplots and Site_Info tables
neon_coords <- Neonplots[ ,2:3] %>%
  select(Latitude = field_latitude, Longitude = field_longitude)
Site_Info_coords <- Site_Info %>%
  select(Site_Number, lat, lon)

# Put Site_Info latitude/longitude in a matrix
Site_Coordinates <- as.matrix(Site_Info[, c("lat", "lon")])

# Same for Neonplots
Neonplots_Coordinates <- as.matrix(Neonplots[, 2:3])

# make list to store matched list
matched <- data.frame(id = numeric(nrow(Neonplots)), # id in forecast
                      site_id = numeric(nrow(Neonplots))) # NEON id

# Loop through each row in Neonplots and calculate each plot's closest site number
for (i in 1:nrow(Neonplots)) {
  
  # Extract the coordinates for the current row in Neonplots
  plot_coords <- Neonplots_Coordinates[i, ]
  
  # Compute distances from this plot to all site coordinates (longlat means its in km)
  distances <- spDistsN1(Site_Coordinates, plot_coords, longlat = TRUE)
  
  # Find the index of the minimum distance for closest site
  closest_site_index <- which.min(distances)
  
  # Assign the corresponding Site_Number from Site_Info to the current row in Neonplots
  matched$id[i] <- Site_Info$Site_Number[closest_site_index]
  matched$site_id[i] <- Neonplots$field_site_id[i]
}

### Get the model output (0715, 16 outputs) (Yang's code)
years <- 2000:2025
Nensem <- 1:25 

# Initialize the output list

sr_list <- list()

# Process Ensemble Function
process_ensemble <- function(ens) {
  ens_id <- sprintf("%05d", ens)
  sr_ens <- list()

  # loop through all the sites
  for (i in seq_len(nrow(matched))) {
    site_id <- matched$site_id[i]
    ID <- matched$id[i]
    
    for (year in years) {
      file_path <- paste0(
        "/projectnb/dietzelab/dongchen/anchorSites/NA_runs/SDA_25ens_2025_4_19/out/ENS-",
        ens_id, "-", ID, "/", year, ".nc"
      )
      
      # open NetCDF files
      nc_data <- tryCatch({
        nc_open(file_path)
      }, error = function(e) {
        return(NULL)
      })
      
      # file open
      if (!is.null(nc_data)) {

        if ("time" %in% names(nc_data$dim)) {
          time_values <- ncvar_get(nc_data, "time")
          time_units <- ncatt_get(nc_data, "time", "units")$value
          
          # Analyze the model start date
          start_datetime <- as.POSIXct(
            sub("days since ", "", time_units),
            format = "%Y-%m-%d %H:%M:%S", tz = "UTC"
          )
          
          # Calculate the time series
          time_datetimes <- as.Date(start_datetime) + (time_values - 1)
          
          # Extract Soil Respiration data
          sr <- tryCatch({
            ncvar_get(nc_data, "SoilResp")
          }, error = function(e) {
            return(NULL)
          })
          
          
          if (!is.null(sr)) {

            sr[sr == -999] <- NA
            
            # create dataframe
            sr_df <- data.table(
              Time = time_datetimes,
              SR = sr,
              Ensemble = ens,
              Site_ID = site_id
            )
            
            # save the results
            sr_ens[[length(sr_ens) + 1]] <- sr_df
          } 
        } 
        
        # Close NetCDF File
        nc_close(nc_data)
      }
    } 
  }
  
  # Combine the results
  sr_result <- if (length(sr_ens) > 0) rbindlist(sr_ens, use.names = TRUE, fill = TRUE) else data.table()
  
  return(list(sr = sr_result))
}


# Apply parallel calculation
num_cores <- detectCores() - 1
results <- mclapply(Nensem, process_ensemble, mc.cores = num_cores)

# Combine results from all ensmebles
sr_output <- rbindlist(lapply(results, `[[`, "sr"))

# Calculate ensemble mean
sr_mean <- sr_output[, .(SR = mean(SR, na.rm = TRUE)), by = .(Time, Site_ID)]

# Merge SR mean value
daily_mean_data <- as.data.table(sr_mean)

# Format Time to character for processing
daily_mean_data[, Time := as.character(Time)]

# Calculate daily SR(time resolution is 1 day)
final_daily_sr_data <- daily_mean_data[, .(
  Final_SR = mean(SR, na.rm = TRUE)), by = .(Time, Site_ID)]

```

Get John's NEON soil respiration data
```{r}
# function to get soil fluxes for specified years for 1 site
soil_flux <- function(site_number, flux_type){
  summary_table <- data.frame(Time = character(5114),
                              Site_id = character(5114),
                              sr_neon = numeric(5114))
  
  day_count <- 1 
  temp <- 0 
  na_count <- 0
  tot <- 0

  years = 2012:2025
  site = matched$site_id[site_number]

  for (year in years){
    for (i in 01:12){
      if (i < 10){
        file_name <- paste0("/projectnb/dietzelab/jzobitz/02-NEON-sites/flux-results/out-flux-", site, "-",year,"-0",i,".Rda")
      }else{
        file_name <- paste0("/projectnb/dietzelab/jzobitz/02-NEON-sites/flux-results/out-flux-", site, "-",year,"-",i,".Rda")
      }
      if (file.exists(file_name)== TRUE){
        load(file_name)
        
        # sort by date so I can average over all horizontal positions
        out_fluxes <- out_fluxes[order(out_fluxes$startDateTime),]
        
        # Average over each day (data is in 30 min increments)
        for (j in 1:(nrow(out_fluxes)/240)){
          if (j == 1){
            start <- 1
            end <- 240
          } else{
            start <- (j-1)*240
            end <- j*240
          }
          for (k in start:end){
            temp <- out_fluxes[[3]][[k]]$flux[flux_type]
            if (is.na(temp)){
              na_count <- na_count + 1}
            else{
              tot <- tot + temp
            }
          }
          summary_table$sr_neon[day_count] <- mean()
          summary_table$Site_id[day_count] <- site
          if (start == 1){
            summary_table$Time[day_count] <- str_sub(out_fluxes$startDateTime[start], end=-1)
          } else{
            summary_table$Time[day_count] <- str_sub(out_fluxes$startDateTime[k], end=-10)
          }
          day_count <- day_count + 1
          na_count <- 0
          tot <- 0
        }
      }
    }
  }
  
return(list(sr = summary_table))
}
r <- soil_flux(1,4)

# Apply parallel calculation for each site
n = 1:nrow(matched)
num_cores <- detectCores() - 1

# Use Marshall method at top of soil (000)
results_j <- mclapply(n, soil_flux, mc.cores = num_cores, flux_type = 4)
# Combine results from all ensembles
sr_marshall <- rbindlist(lapply(results_j, `[[`, "sr"))
# delete rows with all zeros
sr_marshall_filtered <- filter(sr_marshall, sr_neon != 0)

# Use Milington- Quirk method at top of soil (000)
results_j_2 <- mclapply(n, soil_flux, mc.cores = num_cores, flux_type = 8)
# Combine results from all ensembles
sr_m_q <- rbindlist(lapply(results_j_2, `[[`, "sr"))
# delete rows where the soil respiration is 0
sr_m_q_filtered <-filter(sr_m_q, sr_neon != 0)


# combine the two methods
daily_mean_data_neon <- merge(sr_marshall_filtered, sr_m_q_filtered, by = c("Time", "Site_id"))

# rename rows
names(daily_mean_data_neon)[names(daily_mean_data_neon) == "sr_neon.x"] <- "Marshall_SR"
names(daily_mean_data_neon)[names(daily_mean_data_neon) == "sr_neon.y"] <- "Mill_Quirk_SR"
names(daily_mean_data_neon)[names(daily_mean_data_neon) == "Site_id"] <- "Site_ID"


# combine data frame into final_daily_sr_data
daily_sr <- merge(final_daily_sr_data, daily_mean_data_neon, by = c("Time","Site_ID"))
names(daily_sr)[names(daily_sr) == "Final_SR"] <- "Forecast_SR"

# convert Time to date class
daily_sr$Time <- as.Date(daily_sr$Time)

# convert NEON data to to kg/mol ^2 *s (currently in umol/m^2 * s)
daily_sr$Marshall_SR = daily_sr$Marshall_SR*10^-6*12.011*10^-3
daily_sr$Mill_Quirk_SR = daily_sr$Mill_Quirk_SR*10^-6*12.011*10^-3

```

Find error
```{r}
# calculate error
daily_sr$Marsh_error <- daily_sr$Forecast_SR - daily_sr$Marshall_SR
daily_sr$MQ_error <- daily_sr$Forecast_SR - daily_sr$Mill_Quirk_SR

# keep a permanent daily_sr
daily_sr_perm <- daily_sr

# make table for summary statistics
summary_stat <- data.frame(Site_ID = character(length(unique(daily_sr$Site_ID))),
                           Marsh_corr_coef = numeric(length(unique(daily_sr$Site_ID))),
                           MQ_corr_coef = numeric(length(unique(daily_sr$Site_ID))),
                           Marsh_rms_error = numeric(length(unique(daily_sr$Site_ID))),
                           MQ_rms_error = numeric(length(unique(daily_sr$Site_ID))))


# do summary stats
sites <- unique(daily_sr$Site_ID)

for (i in 1:length(sites)){
  temp_data <- filter(daily_sr, Site_ID == sites[i])
  # remove na values
  temp_data <- na.omit(temp_data)
  
  # get rid of first row (so i can remove infinite)
  temp_data <- subset(temp_data, select = -c(Site_ID, Time))
  
  # remove inf values
  temp_data <- temp_data[apply(temp_data, 1, function(row) all(is.finite(row))), ]
  
  # add site name
  summary_stat$Site_ID[i] <- sites[i]
  
  # find correlation coeffs
  summary_stat$Marsh_corr_coef[i] <- cor(temp_data$Marshall_SR, temp_data$Forecast_SR)
  summary_stat$MQ_corr_coef[i] <- cor(temp_data$Mill_Quirk_SR, temp_data$Forecast_SR)
  
  # find RMS errors
  summary_stat$Marsh_rms_error[i] <- sqrt(mean((temp_data$Marshall_SR - temp_data$Forecast_SR)^2))
  summary_stat$MQ_rms_error[i] <- sqrt(mean((temp_data$Mill_Quirk_SR - temp_data$Forecast_SR)^2))
}

```

Graph Time Series
```{r}

for (i in 1:length(sites)){
  temp_data <- filter(daily_sr, Site_ID == sites[i])
  
  
print(ggplot(temp_data) +  
  geom_line(aes(x = Time, y = Marshall_SR, group = 1), linewidth = 1, color = "orange") + 
  geom_line(aes(x = Time, y = Forecast_SR,group = 1), linewidth = 1, color = "lightblue") +
  labs(title = sites[i], x = "date", y = "Soil Respiration (kg/ (m^2 s)"))
}

```


Predicted Obervered plots
```{r}
# make predicted- observed plots
for (i in 1:length(sites)){
  temp_data <- filter(daily_sr, Site_ID == sites[i])
  
  print(ggplot(temp_data, aes(x = Forecast_umol, y = Marsh_umol)) + geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = sites[i], x = "Predicted (kg/m^2*s)", y = "Observed (kg/m^2*s)"))

}
```


Cross-correlation coefficients graphs
```{r}

# List of sites with a noticable lag
lag_sites <- c("MLBS", "HARV","SERC","UNDE","ABBY", "GRSM","WREF","SJER")

peak <- numeric(length(lag_sites))
count = 1

for (site in lag_sites){
  temp_data <- filter(daily_sr, Site_ID == site)
  
   # remove na values
  temp_data <- na.omit(temp_data)
  
  name <- temp_data$Site_ID[1]
  # get rid of first row (so i can remove infinite)
  temp_data <- subset(temp_data, select = -c(Site_ID, Time))
  
  # remove inf values
  temp_data <- temp_data[apply(temp_data, 1, function(row) all(is.finite(row))), ]
  
  ccf_data <- ccf(temp_data$Marshall_SR, temp_data$Forecast_SR, lag = 91, plot = TRUE, ylim = range(-1,1))
  
  title(name)
  
  peak[count] <- ccf_data$lag[which.max(ccf_data$acf)]
  count = count + 1
}

```


Soil Moisture Fraction from Model 
```{r}

# Initialize the output list
sm_list <- list()

# Process Ensemble Function

process_ensemble_sm <- function(ens) {
  ens_id <- sprintf("%05d", ens)
  sm_ens <- list()

  # go through all the sites
  for (i in seq_len(nrow(matched))) {
    site_id <- matched$site_id[i]
    ID <- matched$id[i]
    
    # go through the years
    for (year in years) {
      file_path <- paste0(
        "/projectnb/dietzelab/dongchen/anchorSites/NA_runs/SDA_25ens_2024_11_25/run/ENS-",
        ens_id, "-", ID, "/", year, ".nc"
      )
      
      # check if the file exists
      
      # open NetCDF files
      nc_data <- tryCatch({
        nc_open(file_path)
      }, error = function(e) {
        return(NULL)
      })
      
      # file open
      if (!is.null(nc_data)) {
        if ("time" %in% names(nc_data$dim)) {
          time_values <- ncvar_get(nc_data, "time")
          time_units <- ncatt_get(nc_data, "time", "units")$value
          
          # Analyze the model start date
          start_datetime <- as.POSIXct(
            sub("days since ", "", time_units),
            format = "%Y-%m-%d %H:%M:%S", tz = "UTC"
          )
          
          # Calculate the time series
          time_datetimes <- as.Date(start_datetime) + (time_values - 1)
          
          # get Soil Respiration data
          sm <- tryCatch({
            ncvar_get(nc_data, "SoilMoistFrac")
          }, error = function(e) {
            return(NULL)
          })
          
          
          if (!is.null(sm)) {
            # replace -999 to NA
            sm[sm == -999] <- NA
            
            # create dataframe
            sm_df <- data.table(
              Time = time_datetimes,
              SM = sm,
              Ensemble = ens,
              Site_ID = site_id
            )
            
            # save the results
            sm_ens[[length(sm_ens) + 1]] <- sm_df
          } 
        } 
        
        # Close NetCDF File
        nc_close(nc_data)
      }
    } 
  }
  
  # Combine the results
  sm_result <- if (length(sm_ens) > 0) rbindlist(sm_ens, use.names = TRUE, fill = TRUE) else data.table()
  
  return(list(sm = sm_result))
}

Nensem <- 1:100
# Apply parallel calculation
num_cores <- detectCores() - 1

sm_model_results <- mclapply(Nensem, process_ensemble_sm, mc.cores = num_cores)

# Combine Results from all ensembles
sm_output <- rbindlist(lapply(sm_model_results, '[[', "sm"))

# Calculate ensemble mean
sm_mean <- sm_output[, .(sm = mean(sm, na.rm = TRUE)), by = .(Time, Site_ID)]

# Merge SR mean value with a time resolution of 3h (Not 1 day)
daily_sm_data <- as.data.table(sm_mean)

# Format Time to character for processing
daily_sm_data[, Time := as.character(Time)]

# Calculate daily SR(time resolution is 1 day)
final_daily_sm_data <- daily_sm_data[, .(
  Soil_moist_model = mean(sm, na.rm = TRUE)), by = .(Time, Site_ID)]

```


Soil temperature data from Model
```{r}
process_ensemble_st <- function(ens) {
  ens_id <- sprintf("%05d", ens)
  sm_ens <- list()
  c = 1
  
   # create a data frame
  st_df <- data.table(
              Time = character(13*365*nrow(matched)),
              ST = numeric(13*365*nrow(matched)),
              Site_ID = character(13*365*nrow(matched))
            )

  # go through all the sites
  for (i in seq_len(nrow(matched))) {
    
    
    site_id <- matched$site_id[i]
    ID <- matched$id[i]
            
    # get file path 

      file_path <- paste0(
        "/projectnb/dietzelab/dongchen/anchorSites/NA_runs/SDA_25ens_2024_11_25/run/ENS-",
        ens_id, "-", ID, "/README.txt")
       if (file.exists(file_path)== TRUE){
         # read the text file in 
          temp <- readLines(file_path)
          
         # get the met data path
          met_data_path <- substr(temp[11],17,119)
          
          # change path of readme file so that both 2021 are 2024
          met_data_path <- str_replace(met_data_path,"2021","2024")
          met_data_path <- str_replace(met_data_path,"2021","2024")
          
          # open the new file
          if (file.exists(met_data_path)== TRUE){
            temp_table <- read.table(met_data_path)
            temp_table_yr <- temp_table$V2
            
            for (year in 2012:2024){
              temp_table_yr <- filter(temp_table, V2 == year)
              org_date = paste0(year - 1, "-12-31")
              for (day in 1:366){
                temp_table_yr_day <- filter(temp_table, V3 == day)
                st_df$Site_ID[c] = site_id
                st_df$ST[c] = mean(temp_table_yr_day$V7, na.rm = TRUE)
                
                # need to convert date into "year-mo-day format"
                date <- as.Date(day, origin = org_date) 
                st_df$Time[c] <- format(date, "%Y-%m-%d")
                c = c+1
              }
            }
          }
       }
      }
  return(list(st = st_df))
}

results <- process_ensemble_st(1)
# apply parallel computation

results_temp <- mclapply(Nensem, process_ensemble_st, mc.cores = num_cores)

st_output <- rbindlist(lapply(results_temp, `[[`, "st"))

soil_temperature <- filter(st_output, st != 0)


```


Get Temp and Soil Moisture from John's NEON data
```{r}

soil_temp_and_moist <- function(site_number){
  summary_table <- data.frame(Site_ID = character(5114),
                              Time = character(5114),
                              Soil_temp_neon = numeric(5114),
                              Soil_moist_neon = numeric(5114))

  years = 2012:2025
  site = matched$site_id[site_number]
  day_count <- 1 

  for (year in years){
    for (i in 01:12){
      if (i < 10){
        file_name <- paste0("/projectnb/dietzelab/jzobitz/02-NEON-sites/env-data/env-meas-", site, "-",year,"-0",i,".Rda")
      }else{
        file_name <- paste0("/projectnb/dietzelab/jzobitz/02-NEON-sites/env-data/env-meas-", site, "-",year,"-",i,".Rda")
      }
      if (file.exists(file_name)== TRUE){
        load(file_name)
      
        # get just 501 data
        site_data_temp <- filter(site_data[[2]][[3]], verticalPosition == "501")
        site_data_moist <- filter(site_data[[2]][[2]], verticalPosition == "501")
        
        # sort by date
        site_data_temp <- site_data_temp[order(site_data_temp$startDateTime),]
        site_data_moist <- site_data_moist[order(site_data_moist$startDateTime),]
        
        for (j in 1:(nrow(site_data_temp)/240)){
          if (j == 1){
            start <- 1
            end <- 240
          } else{
            start <- (j-1)*240
            end <- j*240
          }
          summary_table$soil_temp[day_count] <- mean(site_data_temp$soilTempMean[start:end], na.rm = TRUE)
          summary_table$soil_moist[day_count] <- mean(site_data_moist$VSWCMean[start:end], na.rm = TRUE)
          summary_table$Site_id[day_count] <- site
          if (start == 1){
            summary_table$Time[day_count] <- str_sub(site_data_temp$startDateTime[start], end=-1)
          } else{
            summary_table$Time[day_count] <- str_sub(site_data_temp$startDateTime[start], end=-10)
          }
        day_count = day_count + 1
        }
        
      }
    }
  }

return(list(stm = summary_table))
}

# Apply parallel calculation for each site
n = 1:nrow(matched)
num_cores <- detectCores() - 1

# soil_temp results
results_tempandmoist <- mclapply(n, soil_temp_and_moist, mc.cores = num_cores)

# Combine results from all ensembles
soil_tm <- rbindlist(lapply(results_tempandmoist, `[[`, "stm"))

# Delete all rows with just zeros
soil_tm <- filter(soil_tm, soil_temp != 0)

# temp is in C, soil moisture is a ratio 

```
Combine the soil moisture data
```{r}
soil_moist_table <- merge(final_daily_sm_data, soil_tm, by = c("Time","Site_ID"))
```

Graph Time series
```{r}

for (i in 1:length(sites)){
  temp_data <- filter(soil_moist_table, Site_ID == sites[i])
  
  
print(ggplot(temp_data) +  
  geom_line(aes(x = Time, y = Soil_moist_neon, group = 1), linewidth = 1, color = "orange") + 
  geom_line(aes(x = Time, y = Soil_moist_model,group = 1), linewidth = 1, color = "lightblue") +
  labs(title = sites[i], x = "date", y = "Soil Moisture Fraction"))
}

```


Cross- Correlation coefficients
```{r}
lag <- data.frame(lag_sites = character(8),
                 half_lag = numeric(8),
                 year_lag = numeric(8))

lag$lag_sites <- c("MLBS", "HARV","SERC","UNDE","ABBY", "GRSM","WREF","SJER")

peak <- numeric(length(lag$lag_sites))
count = 1

for (site in lag$lag_sites){
  temp_data <- filter(soil_moist_table, Site_ID == site)
  
   # remove na values
  temp_data <- na.omit(temp_data)
  
  name <- temp_data$Site_ID[1]
  # get rid of first row (so i can remove infinite)
  temp_data <- subset(temp_data, select = -c(Site_ID, Time))
  
  # remove inf values
  temp_data <- temp_data[apply(temp_data, 1, function(row) all(is.finite(row))), ]
  
  ccf_data <- ccf(temp_data$Soil_moist_neon, temp_data$Soil_moist_model, lag = 183, plot = TRUE, ylim = range(-1,1))
  
  #title(name)
  
  lag$half_lag[count] <- ccf_data$lag[which.max(ccf_data$acf)]

}
```


To do:
- find function for soil moisture and respiration data, and for soil temperature and respiration



