---
title: "PEcAn Multisite Flat-File Workflow Example"
format: html
editor: visual
---

## Introduction

This notebook demonstrates how to set up a multisite workflow in PEcAn that uses a subset of the sites defined in a flat-file (`site_info.csv`), without querying the database.  
You will learn how to read site metadata, filter by group, and create a MultiSettings object for your runs.

Sites stored in the BETY database can be queried by `sitegroup` to retrieve predefined sets of sites. Inside the database a sitegroup is nothing but a name given to a list of siteIDs, so in a flat file context we can do the same thing by adding one or more grouping columns.

## Example `site_info.csv`

```{r}
# For this demo:
site_info <- tibble::tribble(
	~id, ~lat, ~lon,  ~elev, ~site_group,
	101, 40.1, -88.2, 10,    "NEON",
	102, 41.2, -87.9, 1980,  "NEON",
	201, 39.9, -90.0, 925,   "Ameriflux",
	202, 38.5, -89.5, 1108,  "Ameriflux"
)
# In a real workflow:
# site_info <- read.csv("site_info.csv")
```


## Load Required Libraries

```{r}
library(dplyr)
library(PEcAn.settings)
```


> **NOTE:** You must provide a 'settings' object (as an R list) before running the code below.
> Or load from file: `settings <- read.settings("settings.xml")`

###(sample settings object for running the notebook for multisitesettings)

```{r}
settings <- list(
  outdir = "pecan_runs",
  run = list(
    start.date = "2002-01-01",
    end.date = "2002-12-31",
    site = list(),
    inputs = list(
      met = list(
        id = 99000000001
      )
    )
  ),
  model = list(
    id = 1000000022,
    type = "SIPNET"
  ),
  pfts = list(
  )
)
settings <- as.Settings(settings)

```


## Filter by one group

```{r}
# Filter for NEON sites (change "NEON" to your desired group)
selected_sites <- site_info %>% filter(site_group == "NEON")  
print(selected_sites)           ## prints demo output (not required in main code) 
settings_neon <- createMultiSiteSettings(settings, selected_sites) ## populates settings with metadata for each site in the sitegroup
listToXml(settings_neon, tag = "pecan")

```


## Group by multiple characterstics

Say you're developing a new parameterization for the conifer PFT and you want to test it specifically at high elevation. By filtering and reassigning, this can be done without any edits to the saved `site_info.csv`:

> **NOTE:** Before running the code below, clear your R workspace and recreate the 'settings' object.
> If 'settings' is already a MultiSettings object from a previous run, this code will fail.
> Always start with a fresh 'settings' (as a single Settings object) to avoid errors.

```{r}

settings$pfts <- settings$pfts |>
  append(list(
    name="new_conifer",
    posterior.files = "/path/to/new_conifer/post.distns.Rdata"))
neon_hi_elev <- site_info |>
  filter(
    site_group == "NEON",
    elev > 1000) |>
  mutate(site.pft = "new_conifer")
settings_neon <- createMultiSiteSettings(settings, neon_hi_elev)

## Shows settings object before and after "createMultiSiteSettings(settings, selected_sites)"

```{r}
listToXml(settings, tag = "pecan")
listToXml(settings_neon, tag = "pecan")
```
```



## Summary
- You can use a flat-file (site_info.csv) to manage multisite workflows in PEcAn.
- Filter your sites in R, then pass the filtered data frame to createMultiSiteSettings().
- Group your data by multiple characterstics.
- No database queries are needed for site metadata.


