---
title: "PEcAn Multisite Flat-File Workflow Example"
format: html
editor: visual
---

## Introduction

This notebook demonstrates how to set up a multisite workflow in PEcAn that uses a subset of the sites defined in a flat-file (`site_info.csv`), without querying the database.  
You will learn how to read site metadata, filter by group, and create a MultiSettings object for your runs.

Sites stored in the BETY database can be queried by `sitegroup` to retrieve predefined sets of sites. Inside the database a sitegroup is nothing but a name given to a list of siteIDs, so in a flat file context we can do the same thing by adding one or more grouping columns.

---
Example `site_info.csv`
Suppose you have a CSV file like this: 

```{r}
# For this demo:
site_info <- tibble::tribble(
	~id, ~lat, ~lon,  ~elev, ~site_group,
	101, 40.1, -88.2, 10,    "NEON",
	102, 41.2, -87.9, 1980,  "NEON",
	201, 39.9, -90.0, 925,   "Ameriflux",
	202, 38.5, -89.5, 1108,  "Ameriflux"
)
# In a real workflow:
# site_info <- read.csv("site_info.csv")
```

---

Load Required Libraries

```{r}
library(dplyr)
# library(PEcAn.settings) # Uncomment if needed
# library(PEcAn.all)      # Uncomment if needed
```

---
important !! 

# NOTE: You must provide a 'settings' object (as an R list) before running the code below.
# Example: settings <- list(run = list(start.date = "2025-01-01", end.date = "2025-12-31"))
# Or load from file: settings <- xmlToList(xmlParse("settings.xml"))
---

Filter by one group

```{r}
# Filter for NEON sites (change "NEON" to your desired group)
selected_sites <- site_info %>% filter(site_group == "NEON")  
print(selected_sites)           ## prints demo output (not required in main code) 
settings <- createMultiSiteSettings(settings, selected_sites) ## populates settings with metadata for each site in the sitegroup

```


---

## Group by multiple characterstics

Say you're developing a new parameterization for the conifer PFT and you want to test it specifically at high elevation. By filtering and reassigning, this can be done without any edits to the saved `site_info.csv`:

```{r}
settings$pfts <- settings$pfts |>
  append(list(
    name="new_conifer",
    posterior.files = "/path/to/new_conifer/post.distns.Rdata"))
neon_hi_elev <- site_info |>
  filter(
    site_group == "NEON",
    elev > 1000) |>
  mutate(site.pft = "new_conifer")
settings_neon <- createMultiSiteSettings(settings, neon_hi_elev)
```

---

Example of how settings will look like before and after the execution of "createMultiSiteSettings()" 

Settings object before the execution of 'createMultiSiteSettings()'

```{r}
# Example: settings object with no site info yet

settings <- list(
  run = list(
    start.date = "2025-01-01",
    end.date = "2025-12-31",
    inputs = list()
    # No site info yet!
  )
)

str(settings)  

```

Here this shows the settings object after the execution of "createMultiSiteSettings()"

```{r}
# Example: settings object with site info

settings <- list(
  run = list(
    list(
      site = list(id = 101, lat = 40.1, lon = -88.2, elev = 10, site_group = "NEON", met.start = "2025-01-01", met.end = "2025-12-31"),
      start.date = "2025-01-01",
      end.date = "2025-12-31",
      inputs = list()
    ),
    list(
      site = list(id = 102, lat = 41.2, lon = -87.9, elev = 1980, site_group = "NEON", met.start = "2025-01-01", met.end = "2025-12-31"),
      start.date = "2025-01-01",
      end.date = "2025-12-31",
      inputs = list()
    ),
    list(
      site = list(id = 201, lat = 39.9, lon = -90.0, elev = 925, site_group = "Ameriflux", met.start = "2025-01-01", met.end = "2025-12-31"),
      start.date = "2025-01-01",
      end.date = "2025-12-31",
      inputs = list()
    ),
    list(
      site = list(id = 202, lat = 38.5, lon = -89.5, elev = 1108, site_group = "Ameriflux", met.start = "2025-01-01", met.end = "2025-12-31"),
      start.date = "2025-01-01",
      end.date = "2025-12-31",
      inputs = list()
    )
  )
)

str(settings)

```

---

## Summary
- You can use a flat-file (site_info.csv) to manage multisite workflows in PEcAn.
- Filter your sites in R, then pass the filtered data frame to createMultiSiteSettings().
- Group your data by multiple characterstics.
- No database queries are needed for site metadata.

---

